---
title: "Economic and public health impact of severe weather events in the US (1950-2011)"
author: "Sayef Ishauqe"
date: "September 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```


## Synopsis

Storm data provided by National weather services, US is the primary data source
for this analysis. The storm data for the period of 1950-2011 has been summerized
by storm categories to aggregate total economic loss, injuries and fatalities. 
The type of storm that poses maximum risk of economic loss and casualities have 
been identified. Also, a graphical summary of occurance of such events and their
impact is provided in this report.

## Data Processing

1. The data has been downloaded from National weather service website.

```{r cache=TRUE}
inputURL<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

temp<-tempfile()
download.file(inputURL,temp)

stormData <- read.csv(bzfile(temp))

unlink(temp)

```

2. Unnecessary data elements have been removed from the dataset
3. The Economic loss has been calcualted for ```PROPDMG``` and ```CORPDMG``` 
     fields 
     - ```PROPDMGEXP``` and ```CORPDMGEXP``` data fields are taken into 
     considertion to calculate total economic loss (i.e. if ```PROPDMGEXP``` is 
     "K", ```PROPDMG``` field has been multiplied by 1000)
     - The ```EXP``` fields have been found to have some ambiguous values. As 
     per documentation, B, M or K have been converted accordingly. 
     - Numbers have been converted as a ```10^n```. Other values i.e. -,? are 
     ignored. Such occurance were not significant.
     - A function has been created to calculate and clean-up

```{r}
stormDataAggregate<-select(stormData, EVTYPE,INJURIES,FATALITIES,
                           PROPDMG,PROPDMGEXP, CROPDMG,CROPDMGEXP )

summary(stormDataAggregate)

unique(stormDataAggregate$PROPDMGEXP)

calculateLoss<-function(dmg,dmgexp=""){
     
     if (dmgexp[1]=="K"||dmgexp[1]=="k") result<-dmg*10^3
     else if (dmgexp[1]=="H"||dmgexp[1]=="h") result<-dmg*10^2
     else if (dmgexp[1]=="M"||dmgexp[1]=="m") result<-dmg*10^6
     else if (dmgexp[1]=="B") result<-dmg*10^9
     else if (is.numeric(dmgexp[1])) result<-dmg*10^as.numeric(dmgexp)
     else result <-dmg
     
     
     result
}


stormDataAggregate<-mutate(stormDataAggregate,
                           ECONOMICLOSS<-calculateLoss(PROPDMG,PROPDMGEXP)+
                                calculateLoss(CROPDMG,CROPDMGEXP))


names(stormDataAggregate)[8]<-"ECONOMICLOSS"

```

4. Storm data were summerized by type and total number of injuries, fatalities,
total economic loss and count of occurance have been calculated.

```{r}
stormDataAggregate<-mutate(stormDataAggregate, EVENTCOUNT=1)


stormDataAggregate<-group_by(stormDataAggregate,EVTYPE)

stormDataAggregate<-summarise(stormDataAggregate,sum(INJURIES),sum(FATALITIES),
                         sum(ECONOMICLOSS),sum(EVENTCOUNT))

names(stormDataAggregate)<-c("EVTYPE", "INJURIES", "FATALITIES",
                         "ECONOMICLOSS", "EVENTCOUNT")

stormDataAggregate<-arrange(stormDataAggregate,desc(FATALITIES),
                            desc(ECONOMICLOSS))

head(stormDataAggregate)
```

5. From summerized storm data, storm with highest fatalities, injuries andd 
economic loss have been reported.


## Results

### Impact on population health

```{r}
stormFatalitiesTop10<-head(stormDataAggregate,10)


totalFatalities<-sum(stormDataAggregate$FATALITIES)

stType<-stormFatalitiesTop10[1,1]

pctFatalities<-round(stormFatalitiesTop10[1,3]/totalFatalities*100,0)

```

In the period 1950-2011, ```r stType``` accounts for ```r pctFatalities``` 
percent fatalities

```{r}
stormFatalitiesTop10t<-transform(stormFatalitiesTop10,type='INJURIES',
                                                  occurance=INJURIES)

stormFatalitiesTop10t<-rbind(stormFatalitiesTop10t,
                             transform(stormFatalitiesTop10,type='FATALITIES',
                                       occurance=FATALITIES))

g<-ggplot(stormFatalitiesTop10t,aes(x=reorder(EVTYPE,FATALITIES+INJURIES),
                                   y=occurance,fill=type))
g<-g+geom_bar(stat="Identity")
g<-g+coord_flip()
g<-g+ scale_fill_discrete(name="Type of Casualities")
g<-g+theme( legend.position=c(.8,.5))
g<-g+labs(x="Number of Casualities", y="Type of Storms")
g<-g+ggtitle(label="Top 10 storm types in US in terms of impact on population health")
print(g)
```


### Economic consequences

```{r}

stormEconomicLossTop10<-head(arrange(stormDataAggregate,desc(ECONOMICLOSS)),10)

stormEconomicLossTop10

totalEconomicLoss<-sum(stormDataAggregate$ECONOMICLOSS)

stType<-stormEconomicLossTop10[1,1]
pctLoss<-round(stormFatalitiesTop10[1,4]/totalEconomicLoss*100,0)

```

In the period 1950-2011,```r stType``` accounts for ```r pctLoss``` percent 
economic impact


```{r}

g<-ggplot(stormEconomicLossTop10,aes(reorder(EVTYPE,(ECONOMICLOSS)),
                                     ECONOMICLOSS,fill="topo.colors(1,alpha=.2)"))
g<-g+geom_bar(stat="Identity")
g<-g+coord_flip()
g<-g+theme( legend.position="none")
g<-g+labs(x="Economic Loss", y="Type of Storms")
g<-g+ggtitle(label="Top 10 storm types in US in terms of Economic impact") 
print(g)

```

### Combined result
In both considerations, Tornado has been the most devastating event in this 
period. However, there are a number of storm types which caused similar economic 
impact (at least in logarithmic scale). But, there impact on public health has 
been considerably less.

```{r}
stormTop20<-head(stormDataAggregate,20)

g<-ggplot(stormTop20,aes(x=log10(ECONOMICLOSS),
                                  y=log10(FATALITIES+INJURIES),
                    size=log10(EVENTCOUNT),col=factor(EVTYPE),alpha=1))
g<-g+geom_point(pch=19)
g<-g+theme( legend.position="none")
g<-g+geom_text(aes(x=log10(ECONOMICLOSS)+.1,
                   y=log10(FATALITIES+INJURIES)-.1,
                   label=EVTYPE,size=2.5))
g<-g+ggtitle(label="Relative comparision of top 20 storm types in US (1950-2011)",
                 subtitle = paste(stormTop20[1,]$EVTYPE, "has higest impact in terms of injuries/fatalities and Economic loss"))
g<-g+labs(x="Economic loss in Logarithmic scale (base 10)",
          y="Total injuries and Fatalities in Logarithmic scale (base 10)")
print(g)
```